### 2017-07-17 draft/unique-linear.tex
polymorphism, state monad

\textbf{Let's finally get down to polymorphism. }
\begin{align*}
  \mathit{id} =
  \LAM x\TVAR x & : \forall\TVAR.\TVAR\to\TVAR
  &&\text{seems fine, just instantiate}
  \\
  \mathit{dup} =
  \LAM x\TVAR x & : \forall\TVAR.\TVAR\to(\TVAR,\TVAR)
\end{align*}
\begin{verbatim}
lambda r.
  let (r1, r2) = dup r 
  in  (r1 { a = 42 }, r2.a)
-- should behave like "write before read"
lambda r. 
      (r  { a = 42 }, r .a)
\end{verbatim}
\begin{align*}
  r : \REC{ a:^\WRITE \NUM} & \vdash \UPD r a{42} &&: \REC{ a:^\PVAR \NUM}
  \\
  r : \REC{ a:^\READ \NUM} & \vdash  r. a &&: \NUM
  \\
  r : \REC{ a:^\READ \NUM} & \vdash (\UPD r a{42}, r. a)  &&: \REC{ a:^\PVAR \NUM} \times \NUM
\end{align*}
Using \textit{dup}
\begin{align*}
  r_1 : \REC{ a:^\WRITE \NUM} & \vdash \UPD r a{42} &&: \REC{ a:^\PVAR \NUM}
  \\
  r_2 : \REC{ a:^\READ \NUM} & \vdash  r. a &&: \NUM
  \\
  r_1 : \REC{ a:^\WRITE \NUM},
  r_2 : \REC{ a:^\READ \NUM} & \vdash  (\UPD r a{42} , r. a) &&: \REC{ a:^\PVAR \NUM} \times \NUM
\end{align*}
\end{document}

>>>>

\textbf{Let's finally get down to polymorphism. }
\begin{align*}
  \mathit{id} =
  \LAM x\TVAR x & : \forall\TVAR.\TVAR\to\TVAR
  &&\text{seems fine, just instantiate}
  \\
  \mathit{dup} =
  \LAM x\TVAR x & : \forall\TVAR.\TVAR\to(\TVAR,\TVAR)
\end{align*}
\begin{verbatim}
lambda r.
  let (r1, r2) = dup r 
  in  (r1 { a = 42 }, r2.a)
-- should behave like "write before read"
lambda r. 
      (r  { a = 42 }, r .a)
\end{verbatim}
\begin{align*}
  r : \REC{ a:^\WRITE \NUM} & \vdash \UPD r a{42} &&: \REC{ a:^\PVAR \NUM}
  \\
  r : \REC{ a:^\READ \NUM} & \vdash  r. a &&: \NUM
  \\
  r : \REC{ a:^\READ \NUM} & \vdash (\UPD r a{42}, r. a)  &&: \REC{ a:^\PVAR \NUM} \times \NUM
\end{align*}
Using \textit{dup}
\begin{align*}
  r_1 : \REC{ a:^\WRITE \NUM} & \vdash \UPD r a{42} &&: \REC{ a:^\PVAR \NUM}
  \\
  r_2 : \REC{ a:^\READ \NUM} & \vdash  r. a &&: \NUM
  \\
  r_1 : \REC{ a:^\WRITE \NUM},
  r_2 : \REC{ a:^\READ \NUM} & \vdash  (\UPD r a{42} , r. a) &&: \REC{ a:^\PVAR \NUM} \times \NUM
\end{align*}
\textbf{Suggestion}
Distiguish positive and negative occurrences of a type variable.
\begin{itemize}
\item If a negative occurrence of a type variable is instantiated with a record type, then its fields are all set to
  \READ.
\item Every instantiation of a positive occurrence of a type variable with a record type is allowed
  to have \textbf{different field permissions}. If a record type ends up with permission $\WRITE$,
  then the corresponding record is copied after instantiation.
\item If there is only one positive occurrence under an affine type constructor, then its permission
  is pushed through.
\end{itemize}
That does not quite solve the problem:
\begin{verbatim}
let (r1, r2) = (r, r)
in  (r1 { a = 42 }, r2.a)
\end{verbatim}
Here the reads and writes to \texttt{r1} and \texttt{r2} get out of order due to aliasing in the
pair.

Practical instance: the state monad. Turns out that the bind cannot be handled satisfactorily by the
suggestion on the table because we don't know that \texttt{s} is linear. Seems like linearity is
useful knowledge for polymorphic arguments.
\begin{verbatim}
type ST s a = s -> (s,a)

-- a: -, +
-- s: -, +
returnST :: a -> ST s a
returnST a = \s -> (s, a)

-- b: -, +
-- a: -, +
-- s: +, -, +, -, -, +
bindST :: ST s a -> (a -> ST s b) -> ST s b
bindST ma f = \s -> let (s1,a) = ma s 
                    in  f s1 a
\end{verbatim}
\end{document}
